---
title: "Comparative analysis microarray data of neurons with and without neurofibriallary tangles from Alzheimer's patients."
author: "Max Van Puyvelde"
date: '2023-12'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Comparative analysis microarray data of neurons with and without neurofibriallary tangles from Alzheimer's patients.

Link to dataset: https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-GEOD-4757

## Packages

In this section, the necessary R packages are installed and loaded for data analysis. Various libraries such as affy, limma, and siggenes are installed and loaded into the R environment.

```{r, echo = FALSE, warning = FALSE}
## Install BiocManager
suppressMessages({
    ## Install BiocManager
    if (!require("BiocManager", quietly = TRUE))
        install.packages("BiocManager", quiet = TRUE)
    BiocManager::install(version = "3.16", quiet = TRUE)

    ## Install packages
    BiocManager::install(c("affy", "arrayQualityMetrics", "ArrayExpress", "limma", "siggenes"), update = FALSE, quiet = TRUE)
})
```

```{r,echo=FALSE}
suppressMessages({
    ## Load packages
    library(affy)
    library(arrayQualityMetrics)
    library(ArrayExpress)
    library(limma)
    library(siggenes)
    library(moments)
    library(png)
    library(jpeg)
    library(GOstats)
})
```

## Loading the data

This part involves loading microarray data related to neurons from Alzheimer's patients and organizing it into an object named Exp.

```{r,warning=FALSE}
Exp <- ReadAffy(celfile.path = "E-GEOD-4757/celfiles",phenoData = 'E-GEOD-4757/E-GEOD-4757.sdrf.txt')
```

```{r}
# Basic view
head(exprs(Exp))
pData(Exp)
```

## Quality Control

The quality of the data, both in its raw and log-transformed forms, is assessed using the arrayQualityMetrics function to ensure data integrity and quality before further analysis.

```{r,echo = FALSE, warning=FALSE}
# on raw data
arrayQualityMetrics(
  Exp, 
  intgroup = colnames(pData(Exp))[2],
  outdir = "QC_rawdata", 
  force = TRUE
)
```



```{r,echo = FALSE, warning=FALSE}
# on log transformed data
arrayQualityMetrics(
  Exp,
  intgroup = colnames(pData(Exp))[2],
  outdir = "QC_rawdata_logtransform",
  force = TRUE,
  do.logtransform = TRUE
)
```


## Preprocessing

The raw microarray data is preprocessed using RMA (Robust Microarray Analysis) to normalize and summarize probe-level data into gene-level data for downstream analysis.


```{r}
# Before normalization
boxplot(Exp)

png("boxplot_pre_norm.png", width = 800, height = 600)
boxplot(Exp)
dev.off()
```

The data shows different variances and means, it needs to be normalized.

```{r}
#RMA (Robust Microarray Analysis):multi-chip analysis method that when selected, adjusts the probe intensity data in the following three steps:
# 1)  Signal values for each chip are adjusted using RMA background correction.
# 2)  Data across all of the chips is normalized using normalization method:
# Quantile: Adjusts the probe intensity values so that the distribution is the same across all arrays.
# 3)  Probe data are summarized into gene level signal values using RMA’s median polish summarization.
ExpRMA <- affy::rma(Exp)
```


```{r}
# After normalization
boxplot(ExpRMA)

png("boxplot_post_norm.png", width = 800, height = 600)
boxplot(Exp)
dev.off()
```

As expected, the data now shows equal variances and means.

## Quality control on preprocessed data

A quality assessment is performed on the preprocessed (RMA) data using the arrayQualityMetrics function to validate the reliability of the processed data.

```{r,echo = FALSE, warning=FALSE}
arrayQualityMetrics(ExpRMA, outdir = "rma", force = TRUE) # RMA produces log-transformed data
```


```{r}
head(exprs(ExpRMA))
```

```{r}
# Add categorical variable
annot <- factor(pData(ExpRMA)[, "Factor.Value..clinical.information."])
pData(ExpRMA)[, "Factor.Value..clinical.information."]
```


## Differential expression

The data is prepared and fitted into a linear model to compare gene expression differences between neurons with and without neurofibrillary tangles in Alzheimer's patients.

```{r}
design <- model.matrix(~0 + annot)
colnames(design) <- c("NORMAL", "TANGLES")
fit <- lmFit(ExpRMA, design)
cont.matrix <- makeContrasts(NvsS = NORMAL - TANGLES, levels = design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
```

## Visualisations

Different visual representations such as histograms, volcano plots, MA plots, etc., are generated to depict the distribution of p-values, expression values, and differentially expressed genes.

```{r}
# Histogram of p-values
hist_data_pvalues <- hist(fit2$p.value, breaks = 20, col = "skyblue", main = "Distribution of p-values", xlab = "p-values")
png("pvalues_histogram.png", width = 800, height = 600, res = 120)
plot(hist_data_pvalues, col = "skyblue", main = "Distribution of p-values", xlab = "p-values")
dev.off()

```

```{r}
# Histogram of expression values
hist_data <- hist(unlist(exprs(ExpRMA)), col = "skyblue", xlab = "Expression Values", ylab = "Frequency", main = "Distribution of Expression Values")
png("expression_values_histogram.png", width = 800, height = 600, res = 120)
plot(hist_data, col = "skyblue", xlab = "Expression Values", ylab = "Frequency", main = "Distribution of Expression Values")
dev.off()
```


```{r}
# Volcano plot
jpeg("Volcanoplot.jpg")
volcanoplot(fit2, col = ifelse(fit2$p.value < 0.05, "red", "blue"))
dev.off()
img <- readJPEG("Volcanoplot.jpg")
grid::grid.raster(img)
```


```{r}
# MA plot
jpeg("MAplot.jpg")
limma::plotMA(fit2)
dev.off()
img <- readJPEG("MAplot.jpg")
grid::grid.raster(img)
```

## Correction for multiple testing

```{r}
# Storing differentially expressed features
# Adjust for multiple testing method: Benjamini-Hochberg (BH) 
LIMMAout <- topTable(fit2, adjust = "BH", number = nrow(exprs(ExpRMA)))
head(LIMMAout)
```

While the raw p-values indicate statistical significance for a large number of features, the adjusted p values show no statistical significance.

```{r}
# Check intensity values for top results of limma output
exprs(ExpRMA)[rownames(exprs(ExpRMA)) %in% rownames(head(LIMMAout, 1)),]
rowMeans(exprs(ExpRMA)[rownames(exprs(ExpRMA)) %in% rownames(head(LIMMAout)), c(1, 6, 7)])

rowMeans(exprs(ExpRMA)[rownames(exprs(ExpRMA)) %in% rownames(head(LIMMAout)), 2:5])
```
These expression values circle around the median, as was seen in the plotted distribution.

## Annotation

Array annotation information is loaded and checked for consistency with the processed data, ensuring all necessary probes are present for downstream analysis.

```{r}
# Load array annotation file
annotation_MA <- read.table("./E-GEOD-4757/A-AFFY-44.adf.txt", header = TRUE, sep = "\t", skip = 20, fill = TRUE)
print(head(annotation_MA))
```

```{r}
# Check if all probes are present in both sets
dim(annotation_MA)
dim(LIMMAout)
sum(annotation_MA$Composite.Element.Name == sort(rownames(LIMMAout)))
```

Both have 54675 probes
All probes are present

```{r}
# Sort LIMMA output alphabetically on probe name
LIMMAout_sorted <- LIMMAout[sort(rownames(LIMMAout), index.return = TRUE)$ix,]
```

```{r}
# Add gene names to LIMMA output
LIMMAout_sorted$gene <- annotation_MA$Composite.Element.Database.Entry.ensembl.
LIMMAout_annot <- LIMMAout_sorted[sort(LIMMAout_sorted$adj.P.Val, index.return = TRUE)$ix,]
```


```{r}
tail(LIMMAout_annot)
empty_genes <- sum(LIMMAout_annot$gene == "")
non_empty_genes <- sum(LIMMAout_annot$gene != "")
cat("Number of empty genes:", empty_genes, "\n")
cat("Number of non-empty genes:", non_empty_genes, "\n")
```

As we can see, a lot of genes are missing in this array annotation but the majority is present.


## Additional visualisation

This section focuses on creating a histogram to visualize the distribution of log-fold change (logFC) values obtained from the differential expression analysis.

```{r}
# Histogram of logFC values
# Create a histogram of logFC values
hist_logFC <- hist(LIMMAout_annot$logFC, breaks = 100, main = "Distribution of logFC values")
png("logFC_histogram.png", width = 800, height = 600, res = 120)
plot(hist_logFC, main = "Distribution of logFC values")
dev.off()
```


## Enrichment analysis

Enrichment analysis using Gene Ontology (GO) is performed to identify any functional enrichments among differentially expressed genes, highlighting terms with significant gene involvement.

```{r}
# Because we have no significantly expressed genes based on the adjusted p values, we will consider the genes that are differentially expressed based on raw pvalues.

raw_p_values <- LIMMAout_annot$P.value
significant_genes <- LIMMAout_annot$gene[raw_p_values < 0.05]
enriched_GO <- goana(as.numeric(LIMMAout_annot$logFC)[LIMMAout_annot$gene %in% significant_genes], ontology = "BP", number = 30)
```
```{r}
# The terms with most "differntially expressed genes"
top_30_N <- enriched_GO[order(enriched_GO$N, decreasing = TRUE), ]
top_30_N <- head(top_30_N, 30)
top_30_N
```

```{r}
# Some more niche "differentially expressed" genes
bottom_30_N <- enriched_GO[order(enriched_GO$N, decreasing = FALSE), ]
bottom_30_N <- head(bottom_30_N, 30)
bottom_30_N
```

The terms with the highest number of genes are obviously very general genes while the lowest 30 are more niche.

DISCLAIMER: Because we found no statistically significant genes after adjusting the pvalues for multiple testing, these results are not meaningful.
